Powering Hidden Markov Model by Neural Network based Generative Models

### 2019-11-17 manuscript/main.tex
minor change
include precision and f1 in experiment section

%\ecaisubmission   % inserts page numbers. Use only for submission of paper.
                  % Do NOT use for camera-ready version of paper.

@@

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybrid. A hierarchical training is carried out by: i) training a HMM first, ii) then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn. \textcolor{red}{explain more here...}

@@

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. GenHMM has the tractability of exact likelihood.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\item Verifying the efficiency of the proposed model on practical sequential data.
\end{itemize}

@@

\section{Experiments}
To show the validity of our model, we implement our model in PyTorch and test it with sequential data. The code is available at github repository\footnote{Due to double-blind review, the code is submitted in supplementary file.}. We first discuss the experimental setups and then show the experimental results.

@@

Our experiments are performed for: i) standard classification tasks (Table~\ref{tab:acc-classification39f_a}, \ref{tab:acc-classification39f_b}, \ref{tab:acc-classification13f_a}, \ref{tab:acc-classification13f_b}), ii) classification under noise perturbation (table~\ref{tab:acc-classification39f_noise_snr}, \ref{tab:acc-classification39f_noise_type}).
In all experiments, generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. Specifically, our generator structure follows that of a RealNVP described in \cite{2016arXiv160508803D}.
As discussed, the coupling layer shown in \eqref{eq-gl} maps a part of its input signal identically.
The implementation is such that layer $l+1$ would alternate the input signal order of layer $l$ such that no signal remains the same after two consecutive coupling layers.
We term such a pair of consecutive coupling layers as a \textit{flow block}.
In our experiments, each generator $\bm{g}_{s,\kappa}$ consists of four \textit{flow blocks}.
The density of samples in the latent space is defined as Normal, i.e. $p_{s,\kappa}(\bm{z})$ is the density function of standard Gaussian.

@@

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and folded $39$ phonemes.}\label{tab:acc-classification39f_a}
  \centering  
  \begin{tabular}{l|c|c|c}
    \toprule
    {Model} &  K=1 &  K=3 &  K=5  \\  \midrule
    GMM-HMM & $62.3\%$ &  $68.0\%$ &  $68.7\%$  \\ \midrule
    GenHMM & $76.7\%$ & $77.7\%$ &  $77.7\%$ \\ 
    \bottomrule                                                                  
  \end{tabular}
\end{table}

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and $61$ phonemes.}\label{tab:acc-classification39f_b}
  \centering  
  \begin{tabular}{l|c|c|c} \toprule
    {Model} &  K=1 &  K=3 &  K=5
    \\ \midrule
    GMM-HMM & $53.6\%$ &  $59.6\%$ & $61.9\%$  \\ \midrule
    GenHMM & $69.5\%$ & $70.6\%$ & $70.7\%$   \\ 
    \bottomrule
  \end{tabular}
  \vspace{0.2cm}
\end{table}
We firstly show the phoneme classification using 39 dimensional MFCC features (MFCC coefficients, deltas, and delta-deltas), to validate one possible usage of our proposed model. Since generative training is carried out in our experiments, GMM-HMM is trained and tested as a reference model in our experiments. Training and testing of GMM-HMM is in the same condition as GenHMMs are trained and tested. Dataset usage for GenMM and GMM-HMM is the same, and number of states for GMM-HMM is the same as that for GenHMM in modeling each phoneme. Apart from setting the reference model, we also run the experiment comparisons with different total number of mixture components.

Table \ref{tab:acc-classification39f_a} and \ref{tab:acc-classification39f_b} shows the results for this experiments, in which we test both the folded $39$-phoneme classification case (the conventional way) in table \ref{tab:acc-classification39f_a} and the $61$-phoneme classification case in table \ref{tab:acc-classification39f_b}. As shown in both $61$-phoneme and $39$-phoneme cases, GenHMM gets significant higher accuracy than GMM-HMM for the same number of mixture components. As our expectation, GenHMM should have better modeling capacity of sequential data since we bring in the neural network based generators into GenHMM, which should be able to represent complex relationship between states of HMM and sequential data. This is some increase of accuracy as total number of mixture components in GenHMM is increased from $K=1$ to $K=5$. The sequential dependency of data is modeled by HMM itself, while each state of HMM can have better representation using a mixture probabilistic model if data represented by the state is multi-mode. Comparing the results in $39$-phoneme and $61$-phoneme cases, GenHMM gets higher accuracy for $39$-phoneme classification than it does for $61$-phoneme classification. The total training dataset size remains the same as $61$ phonemes are folded into $39$ phonemes. There are less training data available per phonemes and more classes to be recognized in the $61$-phoneme case, which makes the task more challenging.

\begin{table}
  \caption{Test accuracy table for $13$ dimensional features and folded $39$ phonemes.}\label{tab:acc-classification13f_a}
  \centering  
  \begin{tabular}{l|c|c|c}
    \toprule
    {Model} &  K=1 &  K=3 &  K=5  \\  \midrule
    GMM-HMM & $48.5\%$ &  $51.2\%$ &  $52.4\%$  \\ \midrule
    GenHMM & $61.1\%$ & $62.1\%$ & $62.1\%$   \\ 
    \bottomrule                                                                  
  \end{tabular}
\end{table}
\begin{table}
  \caption{Test accuracy table for $13$ dimensional features and $61$
    phonemes.}\label{tab:acc-classification13f_b}
  \centering
  \begin{tabular}{l|c|c|c}
    \toprule
    {Model} &  K=1 &  K=3 &  K=5
    \\ \midrule
    GMM-HMM & $37.1\%$ &  $40.6\%$ & $42.2\%$  \\ \midrule
    GenHMM & $50.3\%$ & $51.0\%$ & $52.3\%$   \\ 
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \caption{Test accuracy table of perturbation with white noise ($K=3$, folded $39$ phonomes).}
  \label{tab:acc-classification39f_noise_snr}
  \vspace{-0.1cm}
  \centering
  \begin{tabular}{l|c|c|c|c}
    \toprule
    {SNR} &  15dB &  20dB &  25dB & 30 dB  \\  \midrule
    GMM-HMM & $36.6\%$ &  $44.2\%$ &  $50.8\%$ & $57.1\%$  \\ \midrule
    GenHMM & $52.4\%$ & $62.0\%$ &  $69.7\%$ & $74.3\%$ \\ 
    \bottomrule                                                                  
  \end{tabular}
  \vspace{0.1cm}
\end{table}
\begin{table}
  \caption{Test accuracy table of perturbation by different type of noise (SNR=$20$dB, $K=3$, folded $39$ phonomes).}
  \label{tab:acc-classification39f_noise_type}
  \vspace{-0.1cm}
  \centering
  \begin{tabular}{l|c|c|c|c}
    \toprule
    {Noise Type} &  White &  Pink &  Babble & Volvo  \\  \midrule
    GMM-HMM & $44.2\%$ &  $48.8\%$ &  $57.7\%$ & $66.6\%$  \\ \midrule
    GenHMM & $62.0\%$ & $65.1\%$ &  $70.0\%$ & $75.7\%$ \\ 
    \bottomrule                                                                  
  \end{tabular}
  \vspace{0.3cm}
\end{table}

Apart from standard classification testing, we also test the robustness of our model to noise perturbations. We train GenHMM with $K=3$ by clean TIMIT training data in the case of folded $39$ phonemes with $39$ dimensional features. The testing dataset is perturbed by either the same type of noise with different signal-to-noise ratio (SNR) as shown in table~\ref{tab:acc-classification39f_noise_snr}, or different type of noises with the same SNR as shown in table~\ref{tab:acc-classification39f_noise_type}. The noise data is from NOISEX-92 database. The baseline of these two sets of experiments is the accuracy testing of GenHMM and GMM-HMM on clean testing data in the same experimental condition, where GenHMM has $77.7\%$ and GMM-HMM gets $68.0\%$ as shown in table~\ref{tab:acc-classification39f_a}. It is shown in table~\ref{tab:acc-classification39f_noise_snr} that GMM-HMM's performance degenerates more than GenHMM's performance at the same level of noise perturbation, though the accuracy of both models increases along the increase of SNR. Especially, for SNR=$30$dB, the accuracy of GenHMM drops only about $3\%$ (from $77.7\%$ to $74.3\%$), while GMM-HMM encounters more than $10\%$ decrease (from $68.0\%$ to $57.1\%$) due to the noise perturbation. In table~\ref{tab:acc-classification39f_noise_type}, the SNR remains constant and GenHMM is tested with perturbation of different noise types. It is shown that GenHMM still has higher accuracy at different types of noise perturbations than GMM-HMM. Among these four types of noise, white noise shows most significant impact to GenHMM while the impact of volvo noise is negligible.

>>>>

% \ecaisubmission   % inserts page numbers. Use only for submission of paper.
% Do NOT use for camera-ready version of paper.

@@

Another track of related work is hybrid method of temporal neural network models and HMM. In \cite{liu2019lstmHmmHyb,buys2018bridging,vik2016rnnHmm}, a long short-term memory (LSTM) model/recurrent neural network (RNN) is combined with HMM as hybrid. A hierarchical training is carried out by: i) training a HMM first, ii) then doing modified training of LSTM using trained HMM. This hierarchical training procedure is motivated by the intuition of using LSTM or RNN to fill in the gap where HMM can not learn.

@@

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. GenHMM has the tractability of exact likelihood.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\item Verifying the proposed model on practical sequential data.
\end{itemize}

@@

\section{Experiments}
To show the validity of our model, we implement our model in PyTorch and test it with sequential data. We first discuss the experimental setups and then show the experimental results.

@@

Our experiments are performed for: i) standard classification tasks (Table~\ref{tab:acc-classification39f_a}, \ref{tab:acc-classification39f_b}, \ref{tab:acc-classification13f_a}, \ref{tab:acc-classification13f_b}), ii) classification under noise perturbation (table~\ref{tab:acc-classification39f_noise_snr}, \ref{tab:acc-classification39f_noise_type}). The criterion used to report the results includes accuracy, precision and F1 score.
In all experiments, generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. Specifically, our generator structure follows that of a RealNVP described in \cite{2016arXiv160508803D}.
As discussed, the coupling layer shown in \eqref{eq-gl} maps a part of its input signal identically.
The implementation is such that layer $l+1$ would alternate the input signal order of layer $l$ such that no signal remains the same after two consecutive coupling layers.
We term such a pair of consecutive coupling layers as a \textit{flow block}.
In our experiments, each generator $\bm{g}_{s,\kappa}$ consists of four \textit{flow blocks}.
The density of samples in the latent space is defined as Normal, i.e. $p_{s,\kappa}(\bm{z})$ is the density function of standard Gaussian.

@@

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and folded $39$ phonemes.}\label{tab:acc-classification39f_a}
  \centering  
  \begin{tabular}{l|l|c|c|c}
    \toprule
    {Model} & Criterion &  K=1 &  K=3 &  K=5  \\  \midrule
    \multirow{3}{*}{GMM-HMM}
            & Accuracy    & $62.3\%$ &  $68.0\%$ &  $68.7\%$  \\
            & {Precision} & $67.9\%$ &  $72.6\%$ &  $73.0\%$  \\
            & {F1}        & $63.7\%$ &  $69.1\%$ &  $69.7\%$ \\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy    & $76.7\%$   & $77.7\%$ &  $77.7\%$ \\ 
            & {Precision} & $76.9\%$   & $78.1\%$ &  $78.0\%$ \\
            & {F1}        & $76.1\%$   & $77.1\%$ &  $77.0\%$\\
    \bottomrule                                                                  
  \end{tabular}
\end{table}

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and $61$ phonemes.}\label{tab:acc-classification39f_b}
  \centering  
  \begin{tabular}{l|l|c|c|c} \toprule
    {Model} & Criterion & K=1 &  K=3 &  K=5
    \\ \midrule
    \multirow{4}{*}{GMM-HMM}
            & Accuracy    & $53.6\%$ &  $59.6\%$ & $61.9\%$  \\
            & {Precision} & $59.1\%$ &  $63.9\%$ & $65.7\%$ \\
            & {F1}        & $54.7\%$ &  $60.5\%$ & $62.7\%$\\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy    & $69.5\%$ & $70.6\%$ & $70.7\%$   \\
            & {Precision} & $69.2\%$ & $70.5\%$ & $71.0\%$ \\
            & {F1}        & $68.6\%$ & $69.6\%$ & $69.6\%$\\
    \bottomrule
  \end{tabular}
  \vspace{0.2cm}
\end{table}
We firstly show the phoneme classification using 39 dimensional MFCC features (MFCC coefficients, deltas, and delta-deltas), to validate one possible usage of our proposed model. Since generative training is carried out in our experiments, GMM-HMM is trained and tested as a reference model in our experiments. Training and testing of GMM-HMM is in the same condition as GenHMMs are trained and tested. Dataset usage for GenMM and GMM-HMM is the same, and number of states for GMM-HMM is the same as that for GenHMM in modeling each phoneme. Apart from setting the reference model, we also run the experiment comparisons with different total number of mixture components.

Table \ref{tab:acc-classification39f_a} and \ref{tab:acc-classification39f_b} shows the results for this experiments, in which we test both the folded $39$-phoneme classification case (the conventional way) in table \ref{tab:acc-classification39f_a} and the $61$-phoneme classification case in table \ref{tab:acc-classification39f_b}. As shown in both $61$-phoneme and $39$-phoneme cases, GenHMM gets significant higher accuracy than GMM-HMM for the same number of mixture components. The comparisons with regarding to precision and F1 scores show similar trends and also demonstrate significant improvement of GenHMM's performance. As our expectation, GenHMM should have better modeling capacity of sequential data since we bring in the neural network based generators into GenHMM, which should be able to represent complex relationship between states of HMM and sequential data. This is some increase of accuracy as total number of mixture components in GenHMM is increased from $K=1$ to $K=5$. The sequential dependency of data is modeled by HMM itself, while each state of HMM can have better representation using a mixture probabilistic model if data represented by the state is multi-mode. Comparing the results in $39$-phoneme and $61$-phoneme cases, GenHMM gets higher accuracy for $39$-phoneme classification than it does for $61$-phoneme classification. The total training dataset size remains the same as $61$ phonemes are folded into $39$ phonemes. There are less training data available per phonemes and more classes to be recognized in the $61$-phoneme case, which makes the task more challenging.

\begin{table}
  \caption{Test accuracy table for $13$ dimensional features and folded $39$ phonemes.}\label{tab:acc-classification13f_a}
  \centering  
  \begin{tabular}{l|l|c|c|c}
    \toprule
    {Model} & Criterion & K=1 &  K=3 &  K=5  \\  \midrule
    \multirow{3}{*}{GMM-HMM}
            & Accuracy & $48.5\%$ &  $51.2\%$ &  $52.4\%$  \\
            & Precision& $56.2\%$ &  $58.3\%$ &  $59.5\%$  \\
            & F1       & $50.3\%$ &  $53.0\%$ &  $54.2\%$  \\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy & $61.1\%$ &  $62.1\%$ & $62.1\%$   \\ 
            & Precision& $61.1\%$ &  $61.9\%$ &  $62.1\%$  \\
            & F1       & $59.7\%$ &  $60.7\%$ &  $60.2\%$  \\

    \bottomrule                                                                  
  \end{tabular}
\end{table}

\begin{table}
  \caption{Test accuracy table for $13$ dimensional features and $61$
    phonemes.}\label{tab:acc-classification13f_b}
  \centering
  \begin{tabular}{l|l|c|c|c}
    \toprule
    {Model} & Criterion &  K=1 &  K=3 &  K=5 \\ \midrule
    \multirow{3}{*}{GMM-HMM}
            & Accuracy & $37.1\%$ &  $40.6\%$ & $42.2\%$  \\
            &Precision & $44.6\%$ &  $47.4\%$ & $48.8\%$  \\
            & F1       & $38.8\%$ &  $42.1\%$ & $43.7\%$ \\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy & $50.3\%$ & $50.8\%$ & $52.3\%$   \\
            & Precision& $49.3\%$ & $50.9\%$ & $52.1\%$ \\
            &F1        & $47.8\%$ & $48.3\%$ & $49.3\%$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \caption{Test accuracy table of perturbation with white noise ($K=3$, folded $39$ phonomes).}
  \label{tab:acc-classification39f_noise_snr}
  \vspace{-0.1cm}
  \centering
  \begin{tabular}{l|l|c|c|c|c}
    \toprule
    \multirow{2}{*}{Model} & \multirow{2}{*}{Criterion} &
                              \multicolumn{4}{c}{White Noise SNR} \\
    \cline{3-6}
                          && 15dB &  20dB &  25dB & 30 dB  \\
    \midrule
    \multirow{3}{*}{GMM-HMM}
            & Accuracy & $36.6\%$ &  $44.2\%$ &  $50.8\%$ & $57.1\%$
 \\
            &Precision & $59.2\%$ &  $64.2\%$ &  $68.4\%$ & $70.6\%$  \\
            & F1       & $39.9\%$ &  $47.7\%$ &  $53.9\%$ & $59.9\%$ \\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy & $52.4\%$ & $62.0\%$ &  $69.7\%$ & $74.3\%$ \\
            &Precision & $60.0\%$ &  $65.9\%$ &  $71.7\%$ & $74.8\%$  \\
            & F1       & $52.5\%$ &  $62.0\%$ &  $69.3\%$ & $73.5\%$ \\
    \bottomrule                                                                  
  \end{tabular}
  \vspace{0.1cm}
\end{table}
\begin{table}
  \caption{Test accuracy table of perturbation by different type of noise (SNR=$20$dB, $K=3$, folded $39$ phonomes).}
  \label{tab:acc-classification39f_noise_type}
  \vspace{-0.1cm}
  \centering
  \begin{tabular}{l|l|c|c|c|c}
    \toprule
    \multirow{2}{*}{Model} & \multirow{2}{*}{Criterion} &
                              \multicolumn{4}{c}{Noise Type} \\
    \cline{3-6}

                           &  &  White &  Pink &  Babble & Volvo  \\
    \midrule
    \multirow{3}{*}{GMM-HMM}
            & Accuracy & $44.2\%$ &  $48.8\%$ &  $57.7\%$ & $66.6\%$
    \\
            &Precision & $64.2\%$ &  $66.1\%$ &  $67.0\%$ & $71.9\%$  \\
            & F1       & $47.7\%$ &  $52.3\%$ &  $59.7\%$ & $67.8\%$ \\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy & $62.0\%$ &  $65.1\%$ &  $70.0\%$ & $75.7\%$ \\
            &Precision & $65.9\%$ &  $67.8\%$ &  $70.4\%$ & $75.9\%$  \\
            & F1       & $62.0\%$ &  $64.6\%$ &  $69.0\%$ & $75.3\%$ \\
    \bottomrule                                                                  
  \end{tabular}
  \vspace{0.3cm}
\end{table}

Apart from standard classification testing, we also test the robustness of our model to noise perturbations. We train GenHMM with $K=3$ by clean TIMIT training data in the case of folded $39$ phonemes with $39$ dimensional features. The testing dataset is perturbed by either the same type of noise with different signal-to-noise ratio (SNR) as shown in table~\ref{tab:acc-classification39f_noise_snr}, or different type of noises with the same SNR as shown in table~\ref{tab:acc-classification39f_noise_type}. The noise data is from NOISEX-92 database. The baseline of these two sets of experiments is the accuracy testing of GenHMM and GMM-HMM on clean testing data in the same experimental condition, where GenHMM has $77.7\%$ and GMM-HMM gets $68.0\%$ as shown in table~\ref{tab:acc-classification39f_a}. Similar superior performance of GenHMM with regarding to precision and F1 scores is also shown. It is shown in table~\ref{tab:acc-classification39f_noise_snr} that GMM-HMM's performance degenerates more than GenHMM's performance at the same level of noise perturbation, though the accuracy of both models increases along the increase of SNR. Especially, for SNR=$30$dB, the accuracy of GenHMM drops only about $3\%$ (from $77.7\%$ to $74.3\%$), while GMM-HMM encounters more than $10\%$ decrease (from $68.0\%$ to $57.1\%$) due to the noise perturbation. In table~\ref{tab:acc-classification39f_noise_type}, the SNR remains constant and GenHMM is tested with perturbation of different noise types. It is shown that GenHMM still has higher accuracy at different types of noise perturbations than GMM-HMM. Among these four types of noise, white noise shows most significant impact to GenHMM while the impact of volvo noise is negligible.
